---
title: "Introduction to R"
output:
  rmdformats::downcute:
    downcute_theme: "chaos"
---

# How does R work? 

# Objects and Functions

1.  Almost every expression in R can be conceptualized in terms of creating- or using functions, objects or both.
2.  Technically, functions are objects, but conceptually it is often easier to think of them as separate types of entities.
3.  In general, objects are static pieces of information, which often are associated with a name.
4.  E.g,. `iris` refers to a `data.frame` (data set), and is included with `R`'s `base` package.

```{r}
iris
```

1.  Functions are expressions which takes a set of inputs, and returns an output
2.  Functions can be used to create new objects.
3.  Functions can be identified by the parentheses following the function name.
4.  The parentheses contain the arguments (i.e,. input) of the function.
5.  We'll look closer at functions later on, but here is an example using the `head()` function, to show the first 5 rows of the `iris` dataset.

```{r}
head(iris) # we can see that `head` is a function, since it is followed by parentheses
           # head takes the input/argument `data`, which here is `iris` (contained in the parenthesis)
```

## Objects

1.  In practice objects are containers of information (which often are named)
2.  On a general level we can distinguish between different types of containers and different types of information
3.  Objects are named in R (usually) using the `<-` (and `->`) operator
4.  For most `R` users the most common object is the `data.frame` and `vector` object.

-   A `data.frame` is basically a table with rows and columns
-   A `vector` is a list of values, corresponding to a single column in a `data.frame`

### Example (`vector`):

```{r}
variable_1 <- "John"

"John" -> variable_1

variable_1
```

### Example (`dataframe`):

```{r}
iris
mtcars
```

### Example (functions):

```{r}
head(iris)
```

# `vectors` (variables) and `data.frames` (datasets)

## `vectors`

1.  The most important data type (container) in R, is the vector.
2.  In essence, everything in R is a vector.
3.  A vector is simply a container of different values of a particular type of data (type of information).
4.  For example, datasets consist of multiple columns, where each column is a vector

### Numeric (ratio/scale, depends on use)

1.  Real numbers (i.e., numbers with decimals), correspond to the datatype (type of information) `double` (short for `double floating point number`)

```{r}
numeric_vector <- 26.23

# We can call the object by just writing its name
# If you call the object, and nothing else it will
# in most instances simply print the object

numeric_vector

```

2.  This is a vector of lenght 1, meaning that it is just a single number.
3.  To combine multiple vectors into a single (longer) vector, we can use the `c()` function.

-   `NOTE` the `c` stands for concatenate.

numeric

```{r}
# here we combine 4 vectors (of length 1) into a single vector. 
numeric_vector <- c(1, 2, 3, 4)

# here we combine 2 vectors, one of lenght 1, and one of length 3.
numeric_vector <- c(1, c(2, 3, 4))

```

### Integer (ratio/scale/ordinal, depends on use)

1.  Natural numbers (i.e., whole numbers without decimals)

```{r}
integer_vector <- 12
integer_vector
```

### Strings (nominal)

1.  Characterpatterns stored in " "
2.  This is usually called a string, but in R its technically called a character vector.

```{r}
string_vector <- "John1_-"
string_vector
```

### Factors (nominal)

1.  Basically the same as strings, but is used as a function to create nominal variables from numeric ones

```{r}
# NB factors must (in general) be a vector 
factor_vector <- factor(c(1,2,3,1,2), 
                        levels = c(1,2,3), 
                        labels = c("firstplace", 
                                   "secondplace", 
                                   "thirdplace"))
factor_vector

# NB note that the function is written over multiple lines, this is because 
# the expression is not "finished" on line 1, and R knows it should check the nex line
```

### ordered (ordinal)

1.  The same as factor, but values are ordered
2.  This is achieved by adding "ordered = TRUE"

```{r}
ordered_factor_vector <- factor(c(1,2,3,1,2), 
                                levels = c(1,2,3), 
                                labels = c("firstplace", 
                                           "secondplace", 
                                           "thirdplace"),
                                ordered = TRUE)
ordered_factor_vector
```

### Logical (nominal)

1.  TRUE versus FALSE

```{r}
logical_vector <- FALSE
logical_vector
```

## Subsetting Vectors

1.  Subsetting refers to the process of accessing information in a datatype.
2.  To access information in vectors we can use [[]] and [].
3.  In practice both [] and [[]] do the same thing, where [] can access multiple values at the same time, whilst [[]] only access a single value.
4.  Inside of [] and [[]] we use different types of vectors to select information.
5.  [] takes any length of vector, whilst [[]] only takes a vector of length 1.

```{r}
integer_vector <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
# we can write this more efficiently, using : 
integer_vector <- 1:10

# Subsetting with integers 
# : is commonly used when subsetting vectors. 
# So instead of writing 
integer_vector[c(1, 2, 3)]

# We can write 
integer_vector[1:3]

# We can also subset with logical vectors, these should (usually) be the same length as 
# the vector you are subsetting 
# The > operator, is a logical operator testing if the left hand side is smaller then the right
integer_vector > 3
integer_vector[integer_vector > 3]

# this is the same as writing 
integer_vector[c(FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE)]



```

### Vectorized expressions

1.  One of the most powerful features in `R` is vectorized expressions. In general, when you apply an expression to a vector, the expression to the vector is applied to the vector as a whole.
2.  In many programming langues expressions have to be applied to each element explicitly. In `R` most expressions work the same for a single value (i.e., just a one-length vector) as they would for a vector with a `1000` values.
3.  This is actually why vectors are called vectors.
4.  In essence, (numeric) vectors can be conceptualized as multidimensional numbers, where normal arithmetic is just vector arithmetic for vectors with one dimenstion (1 element).

```{r}
1 + 1                   
#> c(1 + 1)
c(1, 2, 3) + 1          
#> c(1 + 1, 2 + 1, 3 + 1)
c(1, 2, 3) + c(4, 5, 6) 
#> c(1 + 4, 2 + 5, 3 + 6)
c(1, 2, 3) * 2 # element wise multiplication (must not be confused with standard matrix/vector multiplication!)
#> c(1 * 2, 2 * 2, 3 * 2)
c(1, 2, 3) * c(4, 5, 6)
#> c(1 * 4, 2 * 5, 3 * 6)
```

## Dataframes

1.  After vectors, dataframes are (probably) the data type you will encounter most often
2.  Dataframes are essentially just a collection of vectors (stored in columns) with an equal number of rows.
3.  There are multiple ways of accessing data in dataframes.

```{r}
df <- read.csv("ChickWeight.csv") # Here we use the read.csv function to read a csv-file and store it as a named object

# The simplest is to use $, which selects a columns with the corresponding name
df$weight[1:5] 


#NB it is common to denote dataframes as df 

# You can also use [c("variable name", ...)] to select multiple columns
df[c("weight", "X")] 

# You can also use [[]] to select a single columns (as in vectors)
df[["weight"]]

# The same works for integers (as in vectors)
df[c(1,2)]
df[[1]]

```

4.  There are also more complicated ways of subsetting data for dataframes, using logical vectors, and a different notation (matrix-notation).

# Functions

1.  Functions are simply expressions which take an input, and returns an output.
2.  Furthermore, functions often take arguments, which are either specified by position or by named
3.  These are called positional- and named arguments
4.  Functions are always followed by (), where the arguments are given
5.  Arguments are simply pieces of information fed into the function, and can therefore be supplied using objects
6.  Further, outputs from functions can be stored into objects

### Syntax of functions

```{r, eval = FALSE}
function_name(input_1, input_2, ..., input_n) # This is the syntax of a function 

# if you want to store the output of a function you can use the assignment operator <-
output <- function_name(input_1, input_2, ..., input_n)
```

#### Example: mean function

1.  Syntax: `mean(x, trim = 0, na.rm = FALSE, ...)`

Using named arguments

```{r}
weight <- iris$Sepal.Length

mean(x = df$weight, trim = 0, na.rm = TRUE) 
mean(na.rm = TRUE, x = df$weight, trim = 0)

# You can also use some "unique prefix"
# For example trim is the only argument starting with t (or )

# this won't work
# mean(0, df$weight)
mean(tr = 0, df$weight)
```

Using positional arguments

```{r}

mean(df$weight, 0, TRUE) 
mean(df$weight, 0)
# This does not work!
# mean(TRUE, df$weight, 0) 

```

Using both

```{r}
mean(df$weight, na.rm = TRUE)

#NB: If a function does not take an explicit argument, it assumes a default. 

```

Help function Perhaps the most usefull function in R: help() Just write help() and begin writing the name of the function you want to view the documentation of, and autocomplete

```{r}
help("mean")
help(mean) #this also works
```

# Working directory

```{r}
# install.packages("devtools")
# devtools::install_github("ihrke/astatur")
```

1.  To access files, and save files it is common to do so within a specified working directory.
2.  This avoids the use of long path-strings to open a file e.g., (or using the GUI)
3.  Depending on your operating system, version of R-studio, what type of file you are using, etc, you might need to change your working directory to the folder for this session.
4.  You can set your working directory once using the GUI or using setwd()

```{r}

# You will probably have to write something like 
# setwd("C:/Users/slupp/OneDrive/Skrivebord/NTNU/Mehmet/PSY8003/..." # and so on...)

```

```{r}
df <- read.csv("ChickWeight.csv")
```

# Exercises

1.  View the iris dataset using the "View()" function tip: use help("View")
2.  Find the mean of the Sepal.Width variable
3.  Find the sd of the Sepal.Width variable (hint: help(sd))
4.  Find out what the "table()" function does, and use it
5.  Find out what the "summary()" function does, and use it
6.  Find out what the "str()" function does, and use it

# Section 2: Datamanagement

1.  Base R is fully capable of doing most data-management tasks
    1.  I also recommend that you familiarize yourself with base R, using it when you can
2.  However, using packages (i.e., user written functions) can often make life easier
3.  One of the most used packages in R is dplyr
4.  dplyr makes datamangement much easier

### Installing packages

1.  To install a package (from CRAN) you can use the function "install.packages()"
2.  To activate a package (i.e., load it into your global environment) use "library()"

```{r}
# install.packages("dplyr") 
library(dplyr)
```

### The mutate function

1.  The mutate function is used to "mutate" your data set, either creating new, or altering existing variables
2.  The first (positional) argument is your dataframe
3.  The second is value, which is the definiton/formula for your new/altered variable'
4.  Importantly the new variable must be compatible with your existing dataframe.
5.  Basically it must (in most cases) be a vector with the same lengt as your dataset (i.e., sam number of rows)'
6.  It can for example not be a list

NB: Before managing your data, you should make it easy to look at and review your data

```{r}
# Firstly, use the View() function to view the datasaet. 
View(df)

# It can also be useful to have some sort of summary of the data.
# base 
summary(df) 
```

Now that we have a good view of our data we can start making changes.

```{r}

# Firstly, I would like to change the weight from grams to kilo
df2 <- mutate(df, weight_kilo = 
               weight/1000)

# I would also to label diet using the mutate and factor function in combination
# When overwriting an original variable (instead of creating a new one) it may be 
# a good idea to store it in a new object -- i.e., a new dataframe

df_labeled <- mutate(df, Diet =
                       factor(Diet,
                              levels = c(1,2,3,4),
                              labels = c("low_calorie",
                                         "medium_calore", 
                                         "high_calorie",
                                         "very high calorie")
    
                                                 ))

calories <- factor(df$Diet, levels = c(1,2,3,4),
                   labels = c("low_calorie",
                  "medium_calorie",
                  "high_calorie",
                  "very_high_calorie"))

df_labeled <- mutate(df, Diet = calories
                  )




# NB i could write this in a single line of code, but this makes it easier to read the code

# Now lets see if it worked 

View(df_labeled)
```

### Now for something a bit trickier

1.  it is easy to convert a "real" categorical variable to a factor using the factor function
2.  It is more complicated when converting from a continuous one
3.  Lets say we want to convert weight into a categorical variable with "feather weight", "light weight", "medium weigth", and "heavy weight"
4.  Then we can accompany the function with the "case_when()" function (also from dplyr)

```{r}
df_labeled <- mutate(df_labeled, weight_class =
                       case_when(weight <= 100 ~ "featherweight",
                                 weight <= 200 ~ "light weight",
                                 weight <= 300 ~ "medium weight",
                                 weight >= 300 ~ "heavy weight",
                                 .default = NA))

# NB: Note that each logical expression is evaluated in order, so that if i run weigth <= 200 before weight <= 100 there are no chickens left with values below 100, since those values are below 200 as well

# The .default argument specifies what to do with values not satisfying any of the previous conditions. In general this should not be used to handle exceptions, not real values 
                             
```

R logical Operators:

1.  Is equal: ==
2.  Does not equal: !=
3.  Less than: \<
4.  Less than or equal: \<=
5.  Greater than: \>
6.  Greater than or equal: \>=
7.  AND: &
8.  OR: \|

```{r}
# For example we can see what happens if there are missing values in our variable weight

# Here i use mutate an case when to replace odd numbers with NA's

df_labeled <- mutate(df_labeled, weight = 
                       case_when(
                         weight %% 2 == 0 ~ weight,
                         .default = NA
                       ))

# If we rerun the previous code we can see NA's are coded as heavy weight 

df_labeled <- mutate(df_labeled, weight_class=
                       case_when(weight <= 100 ~ "featherweight",
                                 weight <= 200 ~ "light weight",
                                 weight <= 300 ~ "medium weight",
                                 .default = "Heavy weight"))

# we should therefore rewrite our code so that .default handles exceptions, not real values


df_labeled <- mutate(df_labeled, weight_class=
                       case_when(weight <= 100 ~ "featherweight",
                                 weight <= 200 ~ "light weight",
                                 weight <= 300 ~ "medium weight",
                                 weight > 300 ~ "Heavy weight",
                                 .default = NA # this is the default so it is not really necessary
                                 ))

```

### Recoding variables

1.  Recoding variables can be done by using the case_when function, but sometimes we do not wish to recode all variables, just a few
2.  If so, we can use the "replace()" function
3.  Lets say we want to recode "featherweight" into "lightweight"

```{r}
df_labeled <- mutate(df_labeled, weight_class = 
                       replace(weight_class, 
                               weight_class == "featherweight",
                               "light weight"))
```

### Renaming variables

```{r}
df_labeled <- rename(df_labeled, weight_division = weight_class)
```

### Filtering data

#### Selecting variables

```{r}
df_filtered <- select(df, X:Chick)
df_filtered <- select(df, -Chick)
df_filtered <- select(df, c(Chick,
                            X, 
                            Time))

df_filtered <- select(df, -c(X, Chick))
```

#### Selecting cases

lets say we only want to analyse those with diet one or two

```{r}


df_filtered <- filter(df, Diet == 1 | Diet == 2)



```

### Merging datasets

```{r}
df_1 <- data.frame(ID = 10:1,
                   V1 = 1:10)
df_2 <- data.frame(ID = 1:10,
                   V2 = 11:20)

# Merging columns "carelessly", since the cases are not ordered the same, cases arent matched
df_3 <- bind_cols(df_1, df_2) #Note this is a function from dplyr
df_3

df_4 <- full_join(df_1, df_2, by = "ID")
df_4

```

1.  We can also merge datasets using the *merge()* function from dplyr.
2.  It automatically matches observations based on variables with the same name

```{r}
merge(df_1, df_2)
```

# Exercises

1.  Use the mutate function to sum X, weight and diet into a single variable
2.  Use mutate to create a factor of diet
3.  Use mutate to create a factor of weight
4.  Use mutate to create an ordered variable of weight
5.  Use mutate to create four dummy variables for diet
6.  Make a new dataset without Chicks with diet 3
7.  Create two dataframes and merge them by case (i.e., by an identifier)

# Extras

### for loops, while and if/else statements

1.  When you cannot find a simple function that solves a problem for you, you can usually use a combination of for/or while loops and if/else statements in combination with exisiting functions to solve your problem
2.  Here i use a for loop with if/else statements to recode a string-vector wrongly formatted (using "one" instead og "1") to a numerical vector

```{r}
string_vector <- c("one", "two", "three", "four", "other") 

for (i in 1:length(string_vector)){
  if (string_vector[i] == "one"){
    string_vector[i] <- "1"
  }
  else if (string_vector[i] == "two"){
       string_vector[i] <- "2"
  }
  else if (string_vector[i] == "three"){
        string_vector[i] <- "3"
  }
  else if (string_vector[i] == "four"){
    string_vector[i] <- "4"
  }
  else {
    string_vector[i] <- "NA"
  }
}

string_vector
```

1.  Note that that there usually is a function/combination of functions which can achieve the same task.
2.  Which often is both easier to write, and easier to read.
3.  In this case we could instead use both the case_when- and recode (in combination with as.integer/as.numerinc) function from dplyr

case_when:

```{r}
string_vector <- c("one", "two", "three", "four", "other") 

string_vector <- case_when(string_vector == "one" ~ 1,
                           string_vector == "two" ~ 2,
                           string_vector == "three" ~ 3,
                           string_vector == "four" ~ 4)
string_vector
```

recode:

```{r}
string_vector <- c("one", "two", "three", "four", "other") 


string_vector <- recode(string_vector,
                        one = 1, 
                        two = 2, 
                        three = 3, 
                        four = 4) 

string_vector 


```

### The pipe operator `|>` (base) and `%>%` (margrittr)

1.  The pipe operator is used to feed arguments/objects into functions
2.  Since function outputs can be used as arguments you can often get very complicated lines of code

<Example:Lets> say we want the mean of all means in our dataset but only for those who have `diet == 1`

```{r}
mean(apply(subset(df, Diet == 1), 1, mean))

# This could instead be written as 

df |> 
  subset(Diet == 1) |>
  apply(array = , 1, mean) |> 
  mean()

# dplyr (or rather magrittr) has its own pipe operator %>% which has some ease of use applications when using dplyr functions, for example allowing for autocomplete for variable names e.g,. 


# NB the pipe operator usually applies to the first argument, so in this instance we must use an empty named argument at the start 
```
